A simplified dynamic bioenergetic model for coral-algal symbiosis and coral bleaching

### 2016-10-25 JTB_ms/JTB_ms.tex
finish drafting main text

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Develop a detailed module of environmental stress.} Of primary
  interest to coral biologists and ecologists is symbiosis dysfunction
  under environmental stress, resulting in coral ``bleaching''--the loss
  of algal symbionts from the association (Jokiel and Coles 1977).
  Photooxidative stress in \emph{Symbiodinium} is considered a primary
  trigger of bleaching in response to high temperature and/or light
  (Weis 2008), and prolonged or severe bleaching can result in
  mortality, though corals sometimes recover their symbionts. Bleaching
  susceptibility, severity, and recovery may by influenced by
  interacting factors such as heterotrophy and nutrient availability
  (Wooldridge 2014b), and the genetic identity of \emph{Symbiodinium}
  (Glynn et al. 2001). To simulate these bleaching-related phenomena, we
  develop a generalized framework linking overreduction of the
  photosynthetic light reactions to downstream impacts of
  photoinhibition and photodamage.
\item
  \emph{Reduce theoretical and mathematical complexity.} Following the
  logic of Jager, Martin, and Zimmer (2013), we exclude certain features
  of formal DEB models in order to capture behaviors of interest with
  the simplest possible formulation. Here, we present a model without
  reserves, maturity, or reproduction (see Kooijman 2010). This
  formulation restricts the model's scope to the bioenergetics of growth
  and symbiosis dynamics in adult corals (i.e., reproduction, larval
  stages, and metamorphosis are not considered), but greatly reduces
  theoretical complexity and parameter numbers, which is advantageous
  given the relative paucity of data for corals. However, our primary
  motivation for reducing complexity was to increase accessibility and
  applicability for biologists and ecologists without requiring
  significant expertise in DEB theory.
\item
  \emph{Provide well-documented, open-access code.} In order to
  facilitate the continued development and application of theoretical
  modeling tools for coral symbioses, we provide open access to the
  model in the form of detailed, commented code written in the R
  language (R Core Team 2014). With an accessible and modular framework,
  we envision this as a resource for futher development by the
  scientific community to include additional complexity and
  problem-specific components. To widen the audience for this work, we
  chose the R language because it is freely available and in common use
  by biologists and ecologists.
\end{enumerate}

@@

In this dynamic bioenergetic model of coral-algal symbiosis, carbon and
nitrogen are acquired by each partner and used to construct biomass. A
graphical representation of the model is presented in Fig. 1, and each
model flux and parameter are defined in Tables 1 and 2, respectively. We
use C-moles as the unit of biomass for consistency with the rigorous
mass balance of DEB theory: 1 C-mole is equivalent to the amount of
biomass containing 1 mole of Carbon atoms. Host biomass (\(H\)),
symbiont biomass (\(S\)), and prey biomass (\(X\)) have fixed, but
different, molar N:C ratios (Table 2). Carbon and nitrogen are combined
to produce biomass by synthesizing units (SU), which are mathematical
specifications of the formation of a product from two substrates; we use
the parallel complementary formulation of Kooijman (2010) to specify
these fluxes. The two state variables are symbiont biomass and coral
biomass; because resources are acquired proportionally to surface area,
and surface area is proportional to volume (i.e., corals are
``V1-morphs'' in DEB terminology (Kooijman 2010)), biomass increases
exponentially during growth (indeed, corals grow exponentially (Bak
1976)). The specific growth rate and the ratio of symbiont to host
biomass are the responses of interest of the system. Below we describe
the formulation of each flux involved in producing these responses.

@@

\begin{minipage}[b]{0.12\columnwidth}\raggedright\strut

@@

\begin{minipage}[b]{0.26\columnwidth}\raggedright\strut

@@

\begin{minipage}[t]{0.12\columnwidth}\raggedright\strut

@@

\begin{minipage}[t]{0.26\columnwidth}\raggedright\strut

@@

\begin{minipage}[b]{0.09\columnwidth}\raggedright\strut

@@

\begin{minipage}[b]{0.23\columnwidth}\raggedright\strut

@@

\begin{minipage}[t]{0.09\columnwidth}\raggedright\strut

@@

\begin{minipage}[t]{0.23\columnwidth}\raggedright\strut

@@

We then specify the input of CO\textsubscript{2} to photosynthesis as
\(j_{CO_2}\), a flux mediated by the host that encompasses potentially
diverse carbon-concentrating mechanisms (CCMs), including active
transport of bicarbonate, carbonic anhydrase-catalyzed conversion of
bicarbonate to CO\textsubscript{2} to promote diffusion toward the
symbiont (Tansik, Fitt, and Hopkinson 2015), and acidification of the
symbiosome to increase localized CO\textsubscript{2} concentrations
around the symbiont (Barott et al. 2014). Since these mechanisms require
energetic input by the host, the magnitude of CO\textsubscript{2}
delivery is proportional to the host's excess carbon, \(j_{eC}\),
assuming that some of this carbon is respired to energize these CCMs.
This formulation means that the symbiont indirectly ensures its own
CO\textsubscript{2} supply by providing fixed carbon to the host
(Wooldridge 2013), which establishes a positive feedback in the model
that leads to the existence of multiple stable states, which we discuss
below in the context of coral bleaching. The parameter \(k_{CO_2}\)
scales the efficacy of host CCMs, which enables the comparison of
different rates of CO\textsubscript{2} delivery that may characterize
different coral species (Wooldridge 2014a). The input of
CO\textsubscript{2} to the photosynthesis SU is therefore specified as:

@@

\section{Results}\label{results}

\subsection{Qualitative dynamics}\label{qualitative-dynamics}

In a constant environment, the system ultimately reaches a steady state
of exponential growth or decline. Under some conditions, alternate
steady states can be reached depending on initial values of symbiont and
host biomass. The mechanism that produces these alternate steady states
is the positive feedback between carbon-limitation of the host and
CO\textsubscript{2}-limitation of photosynthesis. If the host becomes
carbon-limited, the efficacy of its CCMs are reduced and photosynthesis
becomes progressively more CO\textsubscript{2}-limited. In turn, less
photosynthate is translocated to the host, causing further
carbon-limitation and an ultimate decline in biomass; this steady state
represents a dysfunctional symbiosis. However, if the host is not
carbon-limited, it effectively delivers CO\textsubscript{2} to
photosynthesis and ensures its continued receipt of fixed carbon from
the symbiont; this system is nitrogen-limited and represents a
functional symbiosis with positive exponential growth. Either of these
stable states may be reached under (some sets of) constant environmental
conditions, depending on the initial values of symbiont and host
biomass: if symbiont biomass is initially very low (i.e., a ``bleached''
coral), the system cannot escape the carbon-limited state and cannot
grow. However, if symbiont biomass is initially high (i.e., a healthy
coral), then the system maintains positive growth in a nitrogen-limited
state. For practical purposes, we consider only positive growth steady
states in the ``steady state behavior'' section below. In the subsequent
``dynamic behavior'' section, we explore more thoroughly the conditions
that cause the system to switch between alternate stable states, and
interpret these dynamics in the context of coral bleaching.

\subsection{Steady state behavior}\label{steady-state-behavior}

@@

To analyze qualitative behavior, we ran the model to steady state across
gradients of external irradiance and nutrients (Fig. 2), which revealed
patterns consistent with observed phenomena in corals. Predicted growth
rates are low at low light and DIN (\textasciitilde{}0.01
d\textsuperscript{-1}), and begin increasing as both of these factors
increase (Fig. 2A). Low light limits photosynthetic rates, resulting in
less fixed carbon shared with the host and an associated increase in the
symbiont to host biomass ratio (Fig. 2B). In agreement with this trend
are many observations of negative correlation between irradiance and
symbiont density (Stimson 1997; Brown et al. 1999; Fitt et al. 2000;
Titlyanov et al. 2001). As higher light alleviates light-limitation of
photosynthesis, growth becomes less carbon-limited.

@@

The maximum predicted growth rates of \textasciitilde{}0.1
d\textsuperscript{-1}, occurring between \textasciitilde{}10-20 mol
photons m\textsuperscript{-2} s\textsuperscript{-1} light and
\textasciitilde{}4 µM DIN (Fig. 2A), are comparable to the rate of 0.07
d\textsuperscript{-1} measured by Tanaka et al. (2007) under similar
N-enriched conditions. Predicted growth rates under conditions more
typical of reef environments (\textless{}0.5 µM DIN) are
\textasciitilde{}0.01-0.03 d\textsuperscript{-1}. Observed specific
growth rates, which are typically based on skeletal mass, fall near or
below 0.01 d\textsuperscript{-1} (Osinga et al. 2011; Osinga et al.
2012) (though values as high as 0.025 d\textsuperscript{-1} have been
reported (Schutter et al. 2010)). That predicted growth rates are
slightly higher than most observations is not surprising, as the model
does not account for external ecological factors that may reduce growth,
such as competition, predation, and bioerosion. Furthermore, the model
does not explicitly account for skeletal formation, which may not always
be well correlated with biomass growth, potentially due to differential
substrate-limitation (Anthony 2002). Nevertheless, when rates of
increase in coral tissue biomass have been directly measured over short
periods, values are consistent with model predictions (e.g., 0.07
d\textsuperscript{-1} in Tanaka et al. (2007)), and similar rates (0.04
d\textsuperscript{-1}) have been measured in \emph{Aiptasia diaphana}, a
non-calcifying symbiotic anemone (Armoza-Zvuloni et al. 2014).

@@

While bleaching in response to high light alone has been observed
experimentally (Schutter et al. 2011; Downs et al. 2013), mass coral
bleaching events occur in response to high temperature (Hoegh-Guldberg
1999); thus, it is important to justify our consideration of light as
the primary stressor in the model. In reality, light and temperature
interact synergistically (Coles and Jokiel 1978; Jones et al. 1998), and
in fact, any stressor that disrupts the quenching of light energy may
lead to bleaching (Wooldridge 2010; Baker and Cunning 2015). This is
because the proximate cause of coral bleaching is excess excitation
energy, but the upstream events that lead to this situation may be
diverse: indeed, elevated temperature may inhibit Rubisco functioning
(Jones et al. 1998) and the repair of the D1 protein in photosystem II
(Warner, Fitt, and Schmidt 1999), which reduces the capacity of
photochemical quenching and leads to an excess of light energy. In this
way, elevated temperature serves to reduce the threshold above which
light causes stress to the system (Hoegh-Guldberg 1999); importantly,
light is still the proximate stressor. Therefore, we omitted temperature
from the model to maintain a desired level of simplicity, while still
allowing photophysiological stress and bleaching to be simulated with
biological realism in response to light. Thus, while the model is
currently accurate from a conceptual and mechanistic standpoint, its
value as a predictive tool under scenarios of changing temperature will
require further development.

@@

\subsection{Sensitivity analysis}\label{sensitivity-analysis}

Each default parameter value used in the model (Table 2) is based on
carefully selected relevant literature. The derivations of each of these
default parameter values are described in the Supplementary Information.
Nevertheless, it is important to evaluate the sensitivity of the model
to changes in these parameter values. Moreover, these sensitivity plots
serve to demonstrate the behavior of the dynamical system. We measured
fractional change in steady state values in response to fractional
changes in parameter values, relative to their default values, under
environmental conditions typical of coral reefs.

@@

Overall, relative changes in the steady state of the system were less
than the equivalent relative change in parameter value, indicating that
the system maintains internal stability. However, changes in certain
parameter values have greater impacts: increasing \(j_{Nm}\) or
decreasing \(K_N\) both dramatically increase host growth, demonstrating
the strong nitrogen-limitation that characterizes these symbioses (REF).
The parameterr \(\bar{a}^*\) has a strong impact on S:H biomass ratios,
because this parameter determines the amount of light absorbed by
symbionts, thus low values of this parameter increase light-limitation
of the system. Altering the maximum growth and turnover rates of
symbiont and host biomass also have expected impacts on steady states.
Parameters having to do with photooxidative stress and bleaching have
little impact under these environmental conditions, but of course have
larger impacts under higher light (e.g., Fig. S6).

\subsection{Dynamic behavior}\label{dynamic-behavior}

Therefore, we are highly interested in the conditions under which the
system switches from one state to another, representing the conditions
under which corals ``bleach'' due to progressive, reinforcing
carbon-limitation. Interestingly, this state can be reached in multiple
ways, by the acceleration of photoinhibition and photodamage, or by high
external DIN, which causes the symbiont to share less carbon,
subsequently causing its own CO2-limitation, and progressive
C-limitation of the system. \#\# varying light: consequences of excess
light and light-limitation \#\# varying DIN: role of N-limitation and
consequences of elevated N \#\# elevated nutrients cause system to
become C-limited (photo: CO2 limited; host: fixed-C limited) \#\#
C-limitation increases susceptibility to bleaching Importantly, these
negative impacts of high irradiance only arise when light energy exceeds
the capacity of photochemistry and NPQ, which in turn depends on
CO\textsubscript{2} availability. While previous models framed
photophysiological stress as a fixed response to absolute irradiance
levels (Eynaud, Nisbet, and Muller 2011), the current implementation
considers the dynamic balance of multiple energy sinks in the causation
of stress, which is more consistent with current understanding of
symbiosis dysfunction (Wooldridge 2013), and establishes an important
role of the host in providing CO\textsubscript{2} for symbiont
photosynthesis (Tansik, Fitt, and Hopkinson 2015; Hopkinson, Tansik, and
Fitt 2015). \#\# consequences of a less efficient but more tolerant
symbiont \#\# Rate of symbiont repopulation (Berner et al. 1993) \#\#
Bicarbonate addition promotes coral growth (Marubini and Thake 1999)
\#\# Seasonal change in sym. density is larger when nutrients are
elevated (Stimson 1997). Such a phenomenon was predicted by Fitt et al.
2000. \#\# Death rule: negative growth for a certain amount of time
should be defensible.

@@

@@

>>>>

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Develop a detailed module of environmental stress.} Of primary
  interest to coral biologists and ecologists is symbiosis dysfunction
  under environmental stress, resulting in coral ``bleaching''--the loss
  of algal symbionts from the association (Jokiel and Coles 1977).
  Photooxidative stress in \emph{Symbiodinium} is considered a primary
  trigger of bleaching in response to high temperature and/or light
  (Weis 2008), and prolonged or severe bleaching can result in
  mortality, though corals sometimes recover their symbionts. Bleaching
  susceptibility, severity, and recovery may by influenced by
  interacting factors such as heterotrophy and nutrient availability
  (Wooldridge 2014b), and the genetic identity of \emph{Symbiodinium}
  (Glynn et al. 2001). To simulate these bleaching-related phenomena, we
  develop a generalized framework linking overreduction of the
  photosynthetic light reactions to downstream impacts of
  photoinhibition and photodamage.
\item
  \emph{Reduce theoretical and mathematical complexity.} Following the
  logic of Jager, Martin, and Zimmer (2013), we exclude certain features
  of formal DEB models in order to capture behaviors of interest with
  the simplest possible formulation. Here, we present a model without
  reserves, maturity, or reproduction (see Kooijman 2010). This
  formulation restricts the model's scope to the bioenergetics of growth
  and symbiosis dynamics in adult corals (i.e., reproduction, larval
  stages, and metamorphosis are not considered), but greatly reduces
  theoretical complexity and parameter numbers, which is advantageous
  given the relative paucity of data for corals. However, our primary
  motivation for reducing complexity was to increase accessibility and
  applicability for biologists and ecologists without requiring
  significant expertise in DEB theory.
\item
  \emph{Provide well-documented, open-access code.} In order to
  facilitate the continued development and application of theoretical
  modeling tools for coral symbioses, we provide open access to the
  model in the form of detailed, commented code written in the R
  language (R Core Team 2014). With an accessible and modular framework,
  we envision this model as a resource for futher development by the
  scientific community to include additional complexity and
  problem-specific components. To widen the audience for this work, we
  chose the R language because it is freely available and in common use
  by biologists and ecologists.
\end{enumerate}

@@

In this dynamic bioenergetic model of coral-algal symbiosis, carbon and
nitrogen are acquired by each partner and used to construct biomass. A
graphical representation of the model is presented in Fig. 1, and each
model flux and parameter are defined in Tables 1 and 2, respectively. We
use C-moles as the unit of biomass for consistency with the rigorous
mass balance of DEB theory: 1 C-mole is equivalent to the amount of
biomass containing 1 mole of Carbon atoms. Host biomass (\(H\)),
symbiont biomass (\(S\)), and prey biomass (\(X\)) have fixed, but
different, molar N:C ratios (Table 2). Carbon and nitrogen are combined
to produce biomass by synthesizing units (SU), which are mathematical
specifications of the formation of a product from two substrates; we use
the parallel complementary formulation of Kooijman (2010) to specify
these fluxes. The two state variables are symbiont biomass and coral
biomass; because resources are acquired proportionally to surface area,
and surface area is proportional to volume (i.e., corals are
``V1-morphs'' in DEB terminology (Kooijman 2010)), biomass increases
exponentially during growth (indeed, corals grow exponentially (Bak
1976)).

Environmental stress is implemented in the form of photo-oxidative
stress, which is thought to be the primary trigger of coral bleaching
(Lesser 1997; Weis 2008; Wooldridge 2009). To simulate these events, we
model the absorption and quenching of light energy by photochemistry and
non-photochemical quenching, and the phenomenological responses that
occur when these capacities are overwhelmed (i.e., photoinhibition,
photodamage, and symbiont loss). While bleaching in response to high
light alone has been observed experimentally (Schutter et al. 2011;
Downs et al. 2013), mass coral bleaching events occur in response to
high temperature (Hoegh-Guldberg 1999); thus, it is important to justify
our consideration of light as the primary stressor in the model. In
reality, light and temperature interact synergistically (Coles and
Jokiel 1978; Jones et al. 1998), and in fact, any stressor that disrupts
the quenching of light energy may lead to bleaching (Wooldridge 2010;
Baker and Cunning 2015). This is because the proximate cause of
photo-oxidative stress is excess excitation energy, but the upstream
events that lead to this situation may be diverse: indeed, elevated
temperature may inhibit Rubisco functioning (Jones et al. 1998) and the
repair of the D1 protein in photosystem II (Warner, Fitt, and Schmidt
1999), which reduces the capacity of photochemical quenching and leads
to an excess of light energy. In this way, elevated temperature serves
to reduce the threshold above which light stresses the system
(Hoegh-Guldberg 1999); importantly, light is still the proximate
stressor. Therefore, we omitted temperature from the model to maintain a
desired level of simplicity, while still allowing photophysiological
stress and bleaching to be simulated with biological realism in response
to light.

Below we describe the formulation of each flux in the model.

@@

@@

\begin{minipage}[b]{0.25\columnwidth}\raggedright\strut

@@

@@

\begin{minipage}[t]{0.25\columnwidth}\raggedright\strut

@@

\begin{minipage}[b]{0.25\columnwidth}\raggedright\strut

@@

@@

\begin{minipage}[t]{0.25\columnwidth}\raggedright\strut

@@

@@

We then specify the input of CO\textsubscript{2} to photosynthesis as
\(j_{CO_2}\), a flux mediated by the host that encompasses potentially
diverse carbon-concentrating mechanisms (CCMs), including active
transport of bicarbonate, carbonic anhydrase-catalyzed conversion of
bicarbonate to CO\textsubscript{2} to promote diffusion toward the
symbiont (Tansik, Fitt, and Hopkinson 2015), and acidification of the
symbiosome to increase localized CO\textsubscript{2} concentrations
around the symbiont (Barott et al. 2014). Since these active CCMs
require energetic input by the host, the magnitude of
CO\textsubscript{2} delivery is proportional to the amount of fixed
carbon available to the host after satisfying its growth demands,
\(j_{eC}\), assuming that some of this carbon is respired to energize
these CCMs. This formulation means that the symbiont indirectly ensures
its own CO\textsubscript{2} supply by providing fixed carbon (=energy)
to the host (Wooldridge 2013), which establishes a positive feedback in
the model that leads to the existence of multiple stable states, which
we discuss below in the context of coral bleaching. The parameter
\(k_{CO_2}\) scales the efficacy of host CCMs, which enables the
comparison of different rates of CO\textsubscript{2} delivery that may
characterize different coral species (Wooldridge 2014a). The input of
CO\textsubscript{2} to the photosynthesis SU is therefore specified as:

@@

\section{Steady state behavior}\label{steady-state-behavior}

In a constant environment, the system ultimately reaches a steady state
of exponential growth or decline. However, under some conditions, either
of these outcomes may occur depending on initial values of symbiont and
host biomass, indicating the presence of alternate stable states. The
mechanism that produces these alternate stable states is the positive
feedback between carbon-limitation of the host and
CO\textsubscript{2}-limitation of photosynthesis: if symbiont biomass is
initially very low (i.e., a ``bleached'' coral), the system cannot
escape positively reinforcing carbon-limitation and cannot grow.
However, if symbiont biomass is initially high (i.e., a ``healthy''
coral), then the system maintains positive growth in a nitrogen-limited
state. For practical purposes, this section of the manuscript considers
only positive growth steady states under constant environments; the
subsequent ``dynamic behavior'' section explores how changing
environments may cause the system to switch between alternate stable
states, which we interpret in the context of coral bleaching.

To analyze qualitative patterns in steady state behavior, we ran the
model to steady state across gradients of external irradiance and
nutrients (Fig. 2), which revealed patterns consistent with observed
phenomena in corals. Predicted growth rates are low at low light and DIN
(\textasciitilde{}0.01 d\textsuperscript{-1}), and begin increasing as
both of these factors increase (Fig. 2A). Low light limits
photosynthetic rates, resulting in less fixed carbon shared with the
host and an associated increase in the symbiont to host biomass ratio
(Fig. 2B). In agreement with this trend are many observations of
negative correlation between irradiance and symbiont density (Stimson
1997; Brown et al. 1999; Fitt et al. 2000; Titlyanov et al. 2001). As
higher light alleviates light-limitation of photosynthesis, growth
becomes less carbon-limited.

@@

@@

The maximum predicted growth rates of \textasciitilde{}0.1
d\textsuperscript{-1}, occurring between \textasciitilde{}10-20 mol
photons m\textsuperscript{-2} s\textsuperscript{-1} light and
\textasciitilde{}4 µM DIN (Fig. 2A), are comparable to the rate of 0.07
d\textsuperscript{-1} measured by Tanaka et al. (2007) under similar
N-enriched conditions. Predicted growth rates under conditions more
typical of reef environments (\textless{}0.5 µM DIN) are
\textasciitilde{}0.01-0.03 d\textsuperscript{-1}. Observed specific
growth rates, which are typically based on skeletal mass, fall near or
below 0.01 d\textsuperscript{-1} (Osinga et al. 2011; Osinga et al.
2012), though values as high as 0.025 d\textsuperscript{-1} have been
reported (Schutter et al. 2010). It is not surprising that predicted
growth rates are slightly higher than most observations, as the model
does not account for external ecological factors that may reduce growth,
such as competition, predation, and bioerosion. Furthermore, the model
does not explicitly account for skeletal formation, which may not always
correlate with biomass growth, potentially due to differential
substrate-limitation (Anthony 2002). Nevertheless, when growth of coral
tissue biomass has been directly measured over short periods, values are
consistent with model predictions (e.g., 0.07 d\textsuperscript{-1} in
Tanaka et al. (2007)), and similar rates (0.04 d\textsuperscript{-1})
have been measured in \emph{Aiptasia diaphana}, a non-calcifying
symbiotic anemone (Armoza-Zvuloni et al. 2014).

@@

@@

\section{Sensitivity analysis}\label{sensitivity-analysis}

The values used for each parameter in the model (Table 2) are derived
from relevant literature; these derivations are described in the
Supplementary Information. Here we evaluate the sensitivity of the model
to changes in these parameter values, which also serves to demonstrate
the behavior of the dynamical system. We measured fractional change in
steady state values in response to fractional changes in parameter
values, relative to their default values, under environmental conditions
typical of coral reefs.

@@

Overall, relative changes in the steady state of the system are less
than the equivalent relative change in parameter value, indicating that
the system maintains internal stability. However, changes in certain
parameter values have more significant impacts: increasing \(j_{Nm}\) or
decreasing \(K_N\) both dramatically increase host growth (Fig. 3),
demonstrating the strong nitrogen-limitation that characterizes these
symbioses (REF). The parameter \(\bar{a}^*\) has a strong impact on S:H
biomass ratios (Fig. 3) since this parameter determines the amount of
light absorbed by symbionts, with lower values increasing
light-limitation. Altering maximum growth and turnover rates of symbiont
and host also have expected impacts on steady states (Fig. 3).
Parameters relating to photooxidative stress and bleaching have little
impact under these environmental conditions (Fig. 3), but of course have
larger impacts under higher light (e.g., Fig. S6). Sensitivity analyses
conducted under different combinations of external light and nutrients
are presented in the Supplementary Information (Figs. S3-S7)

\section{Dynamic behavior}\label{dynamic-behavior}

The dynamic behavior of the model demonstrates its power to integrate
multiple environmental forcings simultaneously to produced integrated
responses of the system. The potential applications of this are
wide-ranging, and so the model may be highly useful for generating
hypotheses about the way the system will respond to complex sets of
environmental dynamics. Here we present several scenarios to highlight
the power of the model and demonstrate that it can also reproduce more
complex phenomena that have been observed in corals.

\subsection{Seasonal variability}\label{seasonal-variability}

Symbiont densities and coral growth rates are known to vary seasonally,
representing an integrated response to changes in a suite of
environmental factors. Light in particular is a strong driver of these
trends (Stimson:1997p3837; Brown et al. 1999; Fagoonee et al. 1999; Fitt
et al. 2000), with high light associated with lower symbiont abundance
and reduced tissue biomass. The role of light in driving seasonal
changes in symbiont density was demonstrated nicely by Stimson (1997),
who also found that experimental nutrient-enrichment amplified the
light-driven seasonal oscillation. Using the levels of light and
nutrients from this study as inputs, the model reproduces this observed
interaction among environmental factors (Fig. 4), and also provides the
mechanism: under high light in summer, photoinhibition reduces symbiont
biomass and leads to carbon-limitation of host growth regardless of
nutrient status; when light is reduced in winter, these responses become
constrained instead by nitrogen, and therefore increase more when DIN is
enriched (Fig. S8).

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig4.png}
\caption{Figure 4. Light-driven seasonal dynamics of symbiont abundance
and coral growth. Light input (upper panel) was designed as a sinusoidal
curve with a period of one year, with maximum and minimum values of 40
and 22 mol photons m\textsuperscript{-2} d\textsuperscript{-1}, based on
Stimson (1997). The dynamic behavior of symiont to host biomass ratio
(middle panel) and the specific growth rate of host biomass (lower
panel) show seasonal oscillations that are greater in magnitude under
higher nutrients (1e-5 molN L\textsuperscript{-1}; dashed lines)
relative to lower nutrients (1e-7 molN L\textsuperscript{-1}; solid
lines), consistent with the findings of Stimson (1997). Prey density was
set at 1e-5 CmolX L\textsuperscript{-1}.}
\end{figure}

The prediction of higher growth when light is reduced indicates that
growth is not limited by low light in winter, but is actually reduced by
excess light in summer, consistent with the experimental findings of
Schutter et al. (2011). Seasonal summertime reductions in tissue biomass
have also been well-documented in the field (Fitt et al. 2000), along
with reductions in net photosynthetic capacity (\textbf{???}).
Importantly, while light alone (at these levels) may drive seasonal
dynamics in the ways discussed, temperature fluctuations may attenuate
or even reverse the effect of light as cooler winters depress
metabolism; thus, the relative magnitude of fluctuation in temperature,
light, and other factors may produce wide variability in the direction
and magnitude of seasonal changes in growth and symbiont abundance,
depending on location and microhabitat. Nevertheless, the seasonal
variability predicted here (Fig. 4) is consistent with experimental and
field observations for corals, and demonstrates the power of this model
to predict dynamic behavior that mechanistically integrate the impacts
of multiple environmental drivers.

\subsection{Coral bleaching}\label{coral-bleaching}

Coral bleaching is the stress-induced loss of symbiotic algae from coral
tissues, which can occur in response to a variety of environmental
stressors. In most cases, coral bleaching is thought to begin with
photo-oxidative stress in symbiont photosynthesis (Lesser 1997), which
triggers a cascade of events leading to symbiont expulsion (Weis 2008).
As symbionts are expelled, the host receives less fixed carbon, which
may then compromise its ability to deliver CO\textsubscript{2} to
photosynthesis (Wooldridge 2013). Increasing
CO\textsubscript{2}-limitation for remaining symbionts, along with an
amplified internal light environment due to reduced self-shading
(Enríquez, Méndez, and Iglesias-Prieto 2005), may further exacerbate
photo-oxidative stress and accelerate symbiont expulsion, driving the
coral into a bleached state.

While these positive feedbacks have been discussed previously in the
literature, this is the first attempt to implement and explore their
properties within a dynamical model. Interestingly, these feedbacks lead
to alternate stable states. The `healthy' state is characterized by
nitrogen-limitation of both symbiont and host: under these conditions,
the symbiont translocates sufficient carbon to support host growth and
CCMs, which ensures that photosynthesis does not become
CO\textsubscript{2}-limited. However, if carbon translocation is
disrupted (and light is sufficiently high), then the system is driven
into the `bleached' state by photo-oxidative stress and positively
reinforcing carbon- and CO\textsubscript{2}-limitation. In this context,
coral bleaching can be understood as a transition from one stable state
to another, and bleaching thresholds are sets of environmental
conditions that push a healthy-state coral onto a trajectory leading to
a bleached stable state.

We are highly interested in the conditions under which the system
switches from a healthy to a bleached state, and can use this model as a
tool to explore this dynamic behavior. Most straightforwardly, this
switch occurs when irradiance exceeds the capacity of photochemistry and
NPQ, causing the production of ROS that leads to photoinhibition,
photodamage, and symbiont expulsion, thereby engaging the positive
feedbacks that lead to a bleached state (Fig. 5). However, the positive
feedbacks involved in bleaching are not engaged only in response to high
external irradiance; in fact, they depend on the relative balance of
light energy absorption and quenching, which in turn depends on the
availability of CO\textsubscript{2} for photosynthesis. While previous
models framed photo-oxidative stress as a fixed response to absolute
external irradiance (Eynaud, Nisbet, and Muller 2011), our
implementation considers the dynamic balance of multiple energy sinks in
the causation of stress, which is more consistent with current
understanding of symbiosis dysfunction (Wooldridge 2013), and
establishes a critical role of host CCMs in providing
CO\textsubscript{2} for photosynthesis (Tansik, Fitt, and Hopkinson
2015; Hopkinson, Tansik, and Fitt 2015).

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig5.png}
\caption{Figure 5. Bleaching.}
\end{figure}

The importance of host CCM activity establishes significant interactive
roles for other factors in influencing coral bleaching responses. For
example, simulations of high light stress (Fig. 6) demonstrate that
bleaching can be attenuated by heterotrophic feeding, a phenomenon which
has been observed experimentally (Borell et al. 2008). The mechanism
underlying this prediction is that feeding by the host increases host
CCM activity, which delays the onset of CO\textsubscript{2}-limitation
of photosynthesis and reduces bleaching severity. On the other hand,
elevated nutrients exacerbate bleaching (Fig. 6), since higher symbiont
densities are more susceptible to CO\textsubscript{2}-limitation
(Wooldridge 2009). Several experimental (Cunning and Baker 2013; Vega
Thurber et al. 2014) and correlational studies (Wooldridge and Done
2009) are consistent with this mechanistic link between high nutrients
and bleaching.

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig6.png}
\caption{Figure 6. Bleaching and interactive factors.}
\end{figure}

Since bleaching represents an alternative stable state, it follows that
a bleached coral cannot recover unless the external condition that
induced bleaching is alleviated. In natural settings, this typically
occurs by seasonal declines in temperature and light. However, due to
hysteresis, a coral cannot recover along the same trajectory it followed
during bleaching; indeed, the stressor must be alleviated well
\emph{below} the threshold that initially caused bleaching in order for
the system to recover (Fig. 7). This is because under the same external
conditions, a bleached coral (relative to a healthy one) is
characterized by greater light amplification and weaker CCM activity,
which serve to maintain the bleached state. The magnitude of hysteresis
will depend on external nutrients and prey availability, as well as
model parameters, yet the fundamental concept demonstrates a core
behavior of the system.

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig7.png}
\caption{Figure 7. Hysteresis.}
\end{figure}

The hysteresis demonstrated in Fig. 7 reveals another interesting
behavior of the system: during recovery from bleaching, the S:H biomass
ratio reaches a higher value that `overshoots' the level maintained in
the healthy stable state. In fact, such a phenomenon has been observed
in both experimental (Cunning, Silverstein, and Baker 2015) and field
studies (Kemp et al. 2014) documenting recovery from bleaching, and has
been interpreted as a potential `disequilibrium in host-symbiont
regulation' (Kemp et al. 2014). The model explains this `overshoot' as
an approach to a bifurcation point in the dynamical system: during
recovery from the `bleached' state, host growth is more strongly
carbon-limited than symbiont growth (since the symbiont has priority
access to carbon), so S:H biomass increases. S:H biomass will continue
to increase beyond the alternative steady state levels as long as the
symbiont remains carbon-limited; not until nitrogen becomes more
strongly limiting to symbiont growth (hence forcing the symbiont to
share fixed carbon with the host) will the system `flip' back to a state
of high carbon fixation and overall nitrogen-limitation that
characterize a `healthy' symbiosis. Of course, the magnitude of this
`overshoot' will depend on external nutrient concentrations and the rate
of change in the stressor during recovery, though simulations predict
similar magnitudes to what was observed by Kemp et al. (supp fig?).

\section{Conclusions}\label{conclusions}

The dynamic behavior of the model demonstrates its ability to
mechanistically integrate multiple forcing functions, and the degree to
which the system state depends not only on external forcing, but also
the past history of the system. The model represents a unifying
mechanistic framework that reproduces many phenomena observed in corals,
suggesting it has captured many of the important features of the system
in a relatively simple formulation. Furthermore, it allows the model to
be used to predict responses of the system to multiple dynamic
environmental forcing functions in a way that incorporates all positive
and negative feedbacks. This makes the model potentially highly useful
as a tool for generating hypotheses. make N, L, and X go up and down.
For example, the model demonstrates and provides a mechanistic
explanation for: an increase in symbiont abundance with nutrients (REF),
an increase in symbiont abundance with heterotrophic feeding (refs),
decrease in symbiont abundance with light (ref), optimal growth at
intermediate light, rapid `bleaching' above threshold of high light
(ref) along with positive feedbacks (wooldridge, enriquez), mitigation
of bleaching by heterotrophic feeding, exacerbation of bleaching by
elevated nutrients, higher standing stocks of less productive symbionts,
greater seasonal change in symbiont density with elevated nutrients,
overshoot of symbiont density after recovery from bleaching.

``host control''=\textgreater{}N-limitation. C-limitation=host has lost
control of symbiosis: either bleaches or symbiont ``overgrows'' host
with net negative growth. (when does this occur?)

In this context, nitrogen-limitation represents a scenario in which the
host is `in control' of the system: negative feedbacks stabilize the
symbiont to host biomass ratio and the system maintains positive growth.
Carbon-limitation represents a loss of `control', wherein positive
feedbacks result in a loss of capacity for carbon fixation and positive
growth. Importantly, these `control' mechanisms arise entirely passively
through the feedbacks present in the system.

(Very low light also causes the symbiont to not share any carbon,
resulting in very high S:H ratios and negative growth of the coral,
which we interpret as a dysfunctional (and impossible) condition.)

@@

\hypertarget{ref-Borell:2008p108}{}
Borell, Esther M, Ade Yuliantri, Kai Bischof, and Claudio Richter. 2008.
``The effect of heterotrophy on photosynthesis and tissue composition of
two scleractinian corals under elevated temperature.'' \emph{Journal of
Experimental Marine Biology and Ecology} 364: 116--23.

@@

\hypertarget{ref-Lesser:1997p3587}{}
Lesser, Michael P. 1997. ``Oxidative stress causes coral bleaching
during exposure to elevated temperatures.'' \emph{Coral Reefs} 16
(July): 187--92.
